---
title: "Statistical Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---


```{r setup, include=FALSE}
library(MASS)
library(tidyverse)
library(Hmisc)
library(corrplot)

# set knitr defaults
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE
)

# set theme defaults
theme_set(theme_minimal() + theme(legend.position = "bottom"))

# set color scale defaults
options(
    ggplot2.continuous.colour = "viridis",
    ggplot2.continuous.fill   = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete   = scale_fill_viridis_d
```

## Methodology

Based on the results in exploratory data analysis, it is clear that deaths, injuries, damage and houses destroyed by volcanic eruptions can be affected by factors like earthquake, tsunami, VEI and elevation of the volcano. Because deaths is the only outcome that may be affected by all four factors mentioned above, we'll use deaths as the outcome in our model fitting. We'll utilize descriptive data on deaths from the dataset, and since it is an ordinal categorical variable, we'll use **ordinal logistic regression** to assess the association between predictors and the ordinal outcome.

We'll use the `polr` command from the `MASS` package to estimate an ordered logistic regression model. We also specify `Hess=TRUE` to have the model return the observed information matrix from optimization (called the Hessian) which is used to get standard errors.

### Definition
An ordinal variable is a categorical variable in which the levels have a natural ordering. Let $Y$ be an ordinal outcome with $J$ categories. Then $P(Y≤j)$is the cumulative probability of $Y$ less than or equal to a specific category
$j=1,...J-1$. The odds of being less than or equal a particular category can be defined as
$$\frac{P(Y≤j)}{P(Y>j)}$$
for $j=1,...J-1$ since $P(Y>j) = 0$ and dividing by zero is undefined. The **log odds** is also known as the **logit**, so that
$$log\frac{P(Y≤j)}{P(Y>j)}=logit(P(Y≤j))$$
In R’s polr the ordinal logistic regression model is parameterized as
$$logit(P(Y≤j))=β_{j0}-η_{1}x_{1}-η_{2}x_{2}-...-η_{p}x_{p}$$

### Application

We use the **ordinal logistic regression** to evaluate the effect of several factors on casualties and property losses which are the outcomes we are interested in. Predictors are tsunami, earthquake, VEI and elevation in this model. Our ordinal logistic regression model can be parameterized as:
$$logit(P(Y≤j))=β_{j0}-η_{1}X_{1}-η_{2}X_{2}-η_{3}X_{3}-η_{4}X_{4}$$
where,  

* $β_{j0}$is an intercept and represents the log-odds of $Y≤j$ when all the predictors are at 0 or their reference level. 
* $X_{1}$ is earthquake, $X_{2}$ is tsunami, $X_{3}$ is VEI, $X_{4}$ is elevation.
* Each $-η_{k}$ for $k = 1,2,3,4$ is the log of the odds ratio comparing the odds of $Y≤j$ between individuals who differ by 1-unit in $X_{k}$.

## Correlation Analysis

First, we perform a correlation analysis. If two predictors have strong correlation, we can only keep one of them in our regression model. This careful consideration ensures the avoidance of multicollinearity, thereby enhancing the precision and interpretability of our model's results.

Correlation between the variables is shown by the following heatmap:
```{r}
# Import tidied data
tidy_df = 
  read.csv2("data/volcanic-eruption-final.csv", sep = ",", stringsAsFactors = TRUE) 
  

# Select variables of interest
selected_data <- 
  tidy_df |>
  dplyr::select(flag_earthquake, flag_tsunami, volcanic_explosivity_index, elevation)

cor_matrix <- rcorr(as.matrix(selected_data), type = "spearman")

correlation_matrix <- cor_matrix$r
p_value_matrix <- cor_matrix$P
p_value_matrix[is.na(p_value_matrix)] <- 1
corrplot(correlation_matrix, p.mat = p_value_matrix, method = "color",
         tl.col = "black", tl.srt = 45, tl.offset = 0.5, tl.cex = 0.8,
         sig.level = c(0.001, 0.01, 0.05), pch.cex = 0.9, pch.col = "grey20",
         tl.pos = "lt", insig = "label_sig", type = "upper")
corrplot(correlation_matrix, type = "lower", add = TRUE, method = "number",
         tl.pos = "n", cl.pos = "n")
```

From the results of correlation analysis, we can see that there is a positive correlation between `flag_tsunami` and `flag_earthquake`. However, correlation coefficient (r) is 0.35, indicating only a moderate correlation. Therefore, we'll include both `flag_tsunami` and `flag_earthquake` in our later analysis.   
We also notice that there is a negative correlation between `elevation` and `flag_tsunami`. This is probably because volcanoes under the sea are more likely to cause tsunami.



## Ordinal Logistic Regression Model

### Fitting the Model

Then we can fit the following ordinal logistic regression model:

$$volcano\_deaths{\_}description \sim flag\_earthquake + flag\_tsunami + volcanic\_explosivity\_index + elevation$$

In this model, `volcano_deaths_description` is a four-level variable, with levels "Few (~1 to 50)", "Some (~51 to 100)", "Many (~101 to 1000)", "Very Many (~1001 or more)", coded 1, 2, 3, and 4  respectively, that we will use as our outcome variable. 

```{r}
death_df = 
  tidy_df |>
  mutate(volcano_deaths_description = droplevels(volcano_deaths_description, exclude = "None/No Record")) |>
  mutate(volcano_deaths_description = factor(volcano_deaths_description, levels = c("Few (~1 to 50)", "Some (~51 to 100)", "Many (~101 to 1000)", "Very Many (~1001 or more)"))) |>
  filter(volcano_deaths_description != "None/No Record")

logistic_model <- polr(volcano_deaths_description ~ flag_earthquake + flag_tsunami + volcanic_explosivity_index + elevation, data = death_df, Hess=TRUE)

## view a summary of the model
logistic_model |> broom::tidy() |> knitr::kable()
```


The estimated model can be written as:

$$logit(\hat{P}(Y≤1))=3.2838-0.4678×earthquake-1.0003×tsunami-0.6581×VEI-0.0002×elevation$$
$$logit(\hat{P}(Y≤2))=3.7736-0.4678×earthquake-1.0003×tsunami-0.6581×VEI-0.0002×elevation$$
$$logit(\hat{P}(Y≤3))=5.2650-0.4678×earthquake-1.0003×tsunami-0.6581×VEI-0.0002×elevation$$

The estimate for `elevation` appears relatively small, possibly because the unit of measurement for elevation is in meters (m). If we use kilometers (km) instead, the estimate would likely increase.

### P-value

Some people require a p value to be satisfied. In this situation, comparing the t-value to the conventional normal distribution using a z test is one method of computing a p-value. This is only true when there are infinite degrees of freedom, but large samples can well approximate it; as sample size drops, the sample becomes more biased. This method is simple to implement and is utilized in other software programs like Stata. We compute the p-values and then merge them back with the table after storing the coefficient table.
```{r}
## store table
ctable2 <- coef(summary(logistic_model))

## calculate and store p values
p <- pnorm(abs(ctable2[, "t value"]), lower.tail = FALSE) * 2

## combined table
ctable2 <- cbind(ctable2, "p value" = p)

knitr::kable(ctable2)

```
**Interpretations**:

All the four predictors (`flag_earthquake`, `flag_tsunami`, `volcanic_explosivity_index`, `elevation`) have p-values less than 0.05, indicating that each of them is statistically significant at the 0.05 significance level.

### Confidence Interval

We can also get confidence intervals for the parameter estimates. If the 95% CI does not cross 0, the parameter estimate is statistically significant.
```{r}
ci = confint.default(logistic_model) 

knitr::kable(ci)
```
**Interpretations**:

The CIs for all predictors (`flag_earthquake`, `flag_tsunami`, `volcanic_explosivity_index`, `elevation`) do not include 0. Therefore, the parameter estimates are statistically significant.

### Odds Ratio

Another way to interpret logistic regression models is to convert the coefficients into odds ratios. To get the OR and confidence intervals, we just exponentiate the estimates and confidence intervals.
```{r}
exp(cbind(OR = coef(logistic_model), ci)) |> knitr::kable()
```
**Interpretation**:

* For volcano eruptions accompanied by earthquake, the odds of causing more deaths (i.e., very many, many or some versus few) is 1.60 times that of volcano eruptions that are not accompanied by earthquake, holding constant all other variables.
* For volcano eruptions accompanied by tsunami, the odds of causing more deaths (i.e., very many, many or some versus few) is 2.72 times that of volcano eruptions that are not accompanied by tsunami, holding constant all other variables.
* For every one unit increase in volcanic explosivity index (VEI), the odds of causing more deaths (i.e., very many, many or some versus few) is multiplied 1.93 times (i.e., increases 93%), holding constant all other variables.
* A one-unit (1 meter) increase in elevation is not associated with a change in the odds of causing more deaths, but large increase (e.g. > 1000 meters) in elevation may be associated with higher odds of causing more deaths.
